---
title: "R Coding Exercise"
---



```{r}
## Installing the dslabs Package from a Local File

# Download the appropriate Zip file for the 'dslabs' package.
# The exact file you need depends on your version of R.
# In this example, the file 'dslabs_0.8.0.zip' is used, suitable for the 'r-release' version of R.
# 'repos = NULL' tells R that the package is being installed from a local file rather than a repository.
# 'type = "win.binary"' specifies that the package is a Windows binary file.

install.packages("C:/Users/Joaquin/School/DA - 6833 (Summer 2024)/Joaquin_Ramriez_Portfolio_II/coding-exercise/dslabs_0.8.0.zip", 
                 repos = NULL, 
                 type = "win.binary")
```

```{r}
library(tidyverse)  # Loads tidyverse packages, including dplyr.
library(dslabs)     # Loads the dslabs.

```

```{r}
# This function helps access the documentation for the gapminder dataset and provides detailed information about the dataset structure, variables, and source.
help(gapminder)

# The summary() function provides statistical summary of the dataset.Which includes minimum, first quartile, median, mean, third quartile, and maximum values for each variable.
summary(gapminder)

# The class() function is used to determine the type of object that gapminder is.
class(gapminder)

```



## Processing Data


```{r}
data("gapminder")  # Load the gapminder dataset from the dslabs package.

africadata <- gapminder %>%  # Filter the dataset to include only African countries.
  filter(continent == "Africa")  # The filter() function is used to specifically focus on the African Country.


str(africadata)  # Check the structure of the 'africadata'.


summary(africadata)  # Get summary statistics for the 'africadata'.
```

```{r}

africa_infant_life <- africadata %>%  # Created a new object with only infant_mortality and life_expectancy variables.
  select(infant_mortality, life_expectancy)  # Select only the 'infant_mortality' and 'life_expectancy' columns.

str(africa_infant_life)  # Check the structure of the new 'africa_infant_life'.

summary(africa_infant_life)  # Get summary statistics for the 'africa_infant_life'.
```

```{r}

africa_population_life <- africadata %>%  # Created another new object with only population and life_expectancy variables.
  select(population, life_expectancy)  # Select only the 'population' and 'life_expectancy' columns.

str(africa_population_life)  # Check the structure of the 'africa_population_life'.

summary(africa_population_life)  # Get summary statistics for the 'africa_population_life'.
```





## Analyzing Relationships in African Countries' Data (Plotting)


```{r}

# Plotting life expectancy against infant mortality.
plot(africa_infant_life$infant_mortality, africa_infant_life$life_expectancy,
     xlab = "Infant Mortality", ylab = "Life Expectancy",
     main = "Life Expectancy vs Infant Mortality in African Countries",
     pch = 16, col = "blue")


# Plotting life expectancy against population size (log scale x-axis).
plot(africa_population_life$population, africa_population_life$life_expectancy,
     xlab = "Population Size", ylab = "Life Expectancy",
     main = "Life Expectancy vs Population Size in African Countries",
     pch = 16, col ="red",
     log = "x")  # Setting x-axis to log scale.



```

The streaks observed in the population vs life expectancy.
  - The dataset 'africadata' includes observations from different regions represented by multiple data points.
  - Countries with larger populations tend to have better resources.

The negative correlation between infant mortality and life expectancy is higher. 
  - Infant mortality rates tend to be associated with lower life expectancies.
  - Countries with higher infant mortality rates often face challenges in healthcare, sanitation, and nutrition.

Both visualizations highlight important demographic and health trends in Africans. Presenting both challenges and potential areas for improvement. 






### Year 2000 from African Countries Data (More Data Processing)

```{r}
# Check for missing values (NA) in the 'infant_mortality' variable in africadata.
missing_infant_mortality <- table(is.na(africadata$infant_mortality))
missing_infant_mortality
```

**Filtering Data for the Year 2000**
```{r}
# Identify years with missing data for infant mortality.
years_with_missing <- unique(africadata$year[is.na(africadata$infant_mortality)])
years_with_missing

# Selecting year 2000
year_selected <- 2000

# Filter dataset to include only data for the year 2000.
africadata_2000 <- africadata %>% 
  filter(year == year_selected)

# Display the structure of the new object to verify it contains 51 observations and 9 variables.
str(africadata_2000)

# Get a statistical summary of the new object.
summary(africadata_2000)
```


## Plotting Life Expectancy against Infant Mortality - Year 2000 (More Plotting)
```{r}

# Plotting life expectancy against infant mortality for year 2000.
plot(africadata_2000$infant_mortality, africadata_2000$life_expectancy,
     xlab = "Infant Mortality", ylab = "Life Expectancy",
     main = "Life Expectancy vs Infant Mortality in African Countries (Year 2000)",
     pch = 16, col = "blue")



# Plotting life expectancy against population size (log scale x-axis) for year 2000.
plot(africadata_2000$population, africadata_2000$life_expectancy,
     xlab = "Population Size", ylab = "Life Expectancy",
     main = "Life Expectancy vs Population Size in African Countries (Year 2000)",
     pch = 16, col = "red",
     log = "x")  # Setting x-axis to log scale.

# Applying linear regression model to explore relationships.
model_infant_life <- lm(life_expectancy ~ infant_mortality, data = africadata_2000)
model_pop_life <- lm(life_expectancy ~ population, data = africadata_2000)

# Summarize the linear regression models.
summary(model_infant_life)
summary(model_pop_life)


```


 - We observe a negative correlation between life expectancy and infant mortality, higher infant mortality rates are associated with lower life expectancies.
 
 - The life expectancy and population size, show no clear correlation.
 
 - There is a statistically significant negative relation between infant_mortality and life_expectancy in African countries for the year 2000 pvalue (2.826-08). The r-squared 0.4701, tells us that infant mortality is a strong predictor of life expectancies. 
 
 - There is no statistically significant relation between population and life_expectancy in African countries for the year 2000 pvalue (0.6159). The r-squared (0.005176), is significantly low and does not provide us meaningful information.
 

In conclusio, infant mortality showed a significant negative correlation with life expectancy, population size does not appear to have a significant relationship with life expectancy in African countries based on the data from the year 2000.