---
title: "CDC Data Exercise"
---

## Dataset Overview

The dataset **"Potentially Excess Deaths from the Five Leading Causes of Death"** offers a detailed look at how many deaths in different areas of the U.S. Essentially, it highlights how many deaths occurred and what expected. In addition we are also provided with this following information:

Underlying cause of death is based on the International Classification of Diseases, and are as follows:

-   Heart disease (I00-I09, I11, I13, and I20--I51)

-   Cancer (C00--C97)

-   Unintentional injury (V01--X59 and Y85--Y86)

-   Chronic lower respiratory disease (J40--J47)

-   Stroke (I60--I69)

-   Locality (nonmetropolitan vs. metropolitan).

-   [CDC's Potentially Excess Deaths webpage](https://data.cdc.gov/NCHS/NCHS-Potentially-Excess-Deaths-from-the-Five-Leadi/vdpk-qzpr/about_data)

-   The dataset has 206k rows.

-   13 columns:

    -   Year: The year of the data record - 2005 through 2015.
    -   Cause of Death: Primary cause of death based on ICD-10 codes.
    -   State: U.S. state where deaths occurred.
    -   Locality: Classification as nonmetropolitan or metropolitan.
    -   Observed Deaths: Actual number of deaths.
    -   Expected Deaths: Number of deaths expected based on benchmark rates.
    -   Potentially Excess Deaths: Deaths exceeding expected numbers.
    -   Percent Potentially Excess Deaths: Percentage of deaths exceeding expectations.

## Loading and Cleaning the Data

We first begin by loading and cleaning the dataset.

```{r}
# Load necessary libraries
library(tidyverse)
library(readr)
library(RColorBrewer)
library(dplyr)
library(ggplot2)

# Set the working directory to the project directory
data <- here::here("cdcdata-exercise","NCHS_-_Five_Leading_Causes_of_Death.csv")
# Define the file path and load the data
data <- read_csv("NCHS_-_Five_Leading_Causes_of_Death.csv")

# Tell us the name of the columns 
colnames(data)

#output two newline characters
cat("\n\n")

# Display the structure of the dataset
str(data)

cat("\n\n")

# Check for missing values
colSums(is.na(data))

cat("\n\n")


# Remove rows with NA values
data_cleaned <- data %>%
  drop_na()

# Display the structure of the cleaned dataset
str(data_cleaned)
```

## Visualizations and Exploratory Data Analysis

**1) Summary statistics for variables.**

```{r}
# Summary statistics for data 
summary(data)

cat("\n\n")

# Summary statistics for cleaned data
summary(data_cleaned)

```

**2) Visualization of trends over time for causes of death.**

```{r}
# Define a color palette
colors <- c("Heart Disease" = "red", 
            "Cancer" = "blue", 
            "Chronic Lower Respiratory Disease" = "darkgreen", 
            "Stroke" = "orange", 
            "Unintentional Injury" = "grey")

# Aggregate data by Year and Cause of Death, excluding "United States"
annual_totals <- data_cleaned %>%
  filter(State != "United States") %>%
  group_by(Year, `Cause of Death`) %>%
  summarize(Total_Observed_Deaths = sum(`Observed Deaths`, na.rm = TRUE), .groups = 'drop')

# Create the time series line plot with thicker lines and formatted year labels
ggplot(annual_totals, aes(x = Year, y = Total_Observed_Deaths, color = `Cause of Death`, group = `Cause of Death`)) +
  geom_line(linewidth = 1.2) +  # Use 'linewidth' for line thickness
  geom_point(size = 2) +  # Optionally, adjust point size
  scale_x_continuous(breaks = seq(min(annual_totals$Year), max(annual_totals$Year), by = 1),  # Ensure years are whole numbers
                     labels = scales::label_number(big.mark = "")) +  # Format year labels as whole numbers
  scale_color_manual(values = colors) +  # Apply custom colors
  labs(title = "Deaths Over Time by Cause of Death",
       x = "Year",
       y = "Total Observed Deaths") +  #  y-axis label
  theme_minimal() +
  theme(legend.position = "bottom",  # Adjust legend position if necessary
        legend.title = element_blank(),  # Remove legend title
        legend.key.width = unit(0.5, "cm"),  # Adjust legend key width
        legend.key.height = unit(0.5, "cm"),  # Adjust legend key height
        legend.text = element_text(size = 10))  # Adjust legend text size

 
```

**Causes of Death:**

-   Cancer: This is the leading cause of death according to the graph, and the number of deaths has been slowly increasing over time. There appears to be a slight acceleration in the rate of increase in recent years.

-   Heart Disease: Heart disease is the second leading cause of death. The number of deaths due to heart disease has fluctuated somewhat over the years, but there is a general downward trend from 2008 to 2013. Since then, there has been a slight uptick.

-   Unintentional Injuries: Unintentional injuries are the third leading cause of death according to the graph. The number deaths due to unintentional injuries has been increasing steadily over time.

-   Chronic Lower Respiratory Disease and Stroke: Chronic lower respiratory disease and stroke have the two lowest numbers of deaths depicted in the graph. The number of deaths due to these causes appears to be relatively stable.

**3) Visualization: Causes of Death over time (Percentages).**

```{r}
# Aggregate total observed deaths by Year and Cause of Death, excluding "United States"
annual_totals <- data_cleaned %>%
  filter(State != "United States") %>%
  group_by(Year, `Cause of Death`) %>%
  summarize(Total_Observed_Deaths = sum(`Observed Deaths`, na.rm = TRUE), .groups = 'drop')

# Calculate total deaths by Year
total_deaths_by_year <- annual_totals %>%
  group_by(Year) %>%
  summarize(Total_Deaths = sum(Total_Observed_Deaths, na.rm = TRUE))

# Merge the total deaths with the annual totals to calculate percentages
annual_totals_with_percentages <- annual_totals %>%
  left_join(total_deaths_by_year, by = "Year") %>%
  mutate(Percentage_of_Deaths = (Total_Observed_Deaths / Total_Deaths) * 100)

# Reorder the Cause of Death factor levels based on total percentage
annual_totals_with_percentages <- annual_totals_with_percentages %>%
  mutate(`Cause of Death` = reorder(`Cause of Death`, Percentage_of_Deaths, FUN = sum))

# Create a bar plot with percentages and labels, with reordered bars
ggplot(annual_totals_with_percentages, aes(x = Year, y = Percentage_of_Deaths, fill = `Cause of Death`)) +
  geom_bar(stat = "identity", position = "stack") +  # Stacked bar plot
  geom_text(aes(label = sprintf("%.1f%%", Percentage_of_Deaths), y = Percentage_of_Deaths), 
            position = position_stack(vjust = 0.5), size = 3.5, color = "white") +  # Add percentage labels inside bars
  labs(title = "Percentage of Observed Deaths by Cause of Death Over Time",  # Update title
       x = "Year",
       y = "Percentage of Total Deaths") +
  scale_x_continuous(breaks = seq(min(annual_totals_with_percentages$Year), max(annual_totals_with_percentages$Year), by = 1), 
                     labels = scales::label_number(big.mark = "")) +  # Format year labels
  scale_fill_manual(values = colors) +  # Apply custom colors
  theme_minimal() +
  theme(legend.position = "bottom",  # Adjust legend position if necessary
        legend.title = element_blank(),  # Remove legend title
        legend.key.width = unit(0.5, "cm"),  # Adjust legend key width
        legend.key.height = unit(0.5, "cm"),  # Adjust legend key height
        legend.text = element_text(size = 10))  # Adjust legend text size

```

**Percentage of Deaths Over Time:**

-   Cancer is the leading cause of death in the United States, representing around 40% of total deaths from the top 5 causes.

-   Heart Disease is the second leading cause of death (after cancer). The graph shows a slight decrease in the percentage of deaths due to heart disease over time.

-   Unintentional Injuries the graph shows a slight increase in the percentage of deaths due to unintentional injuries, meaning more deaths have occurred from unintentional causes over time.

-   Chronic Lower Respiratory Disease the graph suggests a slight increase in deaths due to chronic lower respiratory disease.

-   Stroke appears to have the lowest percentage of deaths among the top causes depicted in the graph. The percentage has remained relatively stable over time, with a possible slight decrease.

Overall, the graph shows the percentage of deaths by cause over time. The leading causes of death are cancer and heart disease, but the percentage of deaths due to these causes is slowly decreasing. Deaths due to unintentional injuries show a slight upward trend, while chronic lower respiratory disease and stroke show a stable or slightly increasing percentage of deaths.

**4) Visualization: Causes of death by state.**

```{r}
# Exclude rows where State is "United States" and Total_Observed_Deaths is below 650,000
state_observed_deaths <- data_cleaned %>%
  filter(State != "United States") %>%
  group_by(State) %>%
  summarize(Total_Observed_Deaths = sum(`Observed Deaths`, na.rm = TRUE)) %>%
  filter(Total_Observed_Deaths >= 650000) %>%  # Filter out states with less than 650,000 observed deaths
  arrange(desc(Total_Observed_Deaths))

# Plotting the state-wise distribution of observed deaths excluding United States
ggplot(state_observed_deaths, aes(x = reorder(State, -Total_Observed_Deaths), y = Total_Observed_Deaths)) +
  geom_bar(stat = "identity", fill = "skyblue", width = 0.7) +  # Adjust bar width as needed
  labs(title = "Total Observed Deaths by State (Excluding United States)",
       x = "State",
       y = "Total Observed Deaths") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 7))  # Adjust angle and size

```

**Here are the top ten states with the most deaths according to the graph:**

1.  California
2.  Texas
3.  Florida 4.New York
4.  Pennsylvania
5.  Ohio
6.  Illinois
7.  Michigan
8.  North Carolina
9.  Georgia

Intrigued by this, we will now dive deeper and explore the leading cause of death in these top 10 states. This will help provide and understanding mortality patterns across the country.

```{r}
# Specify the states of interest
states_of_interest <- c("California", "Texas", "Florida", "New York", "Pennsylvania", 
                        "Ohio", "Illinois", "Michigan", "North Carolina", "Georgia")

# Filter data for specified states and calculate percentages by cause of death
state_cause_percentages <- data_cleaned %>%
  filter(State %in% states_of_interest) %>%
  group_by(State, `Cause of Death`) %>%
  summarize(Percentage = sum(`Observed Deaths`) / sum(data_cleaned$`Observed Deaths`) * 100, .groups = 'drop') %>%
  arrange(State, desc(Percentage))  # Arrange by State and descending Percentage

# Plotting separate bar plots for each state showing percentage distribution of causes of death
plots_list <- lapply(states_of_interest, function(state) {
  state_data <- state_cause_percentages %>%
    filter(State == state) %>%
    arrange(desc(Percentage))  # Sort within each state from highest to lowest percentage
  
  ggplot(state_data, aes(x = reorder(`Cause of Death`, -Percentage), y = Percentage, fill = `Cause of Death`)) +
    geom_bar(stat = "identity", width = 0.7) +  # Bar plot
    geom_text(aes(label = sprintf("%.1f%%", Percentage), y = Percentage), 
              position = position_stack(vjust = 0.5), size = 3.5, color = "white") +  # Add percentage labels inside bars
    labs(title = paste("Distribution of Causes of Death in", state),
         x = "Cause of Death",
         y = "Percentage") +
    scale_fill_manual(values = colors) +  # Apply custom colors
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels if needed
          plot.title = element_text(hjust = 0.5))  # Center title horizontally
})

# Print each plot separately
for (i in seq_along(plots_list)) {
  print(plots_list[[i]])
}

```

1.  California

-   Cancer: 2%
-   Heart Disease: 1.4%
-   Unintentional Injury: 0.7%
-   Chronic Lower Respiratory Disease: 0.3%
-   Stroke: 0.3%

2.  Texas

-   Cancer: 1.4%
-   Heart Disease: 1.1%
-   Unintentional Injury: 0.6%
-   Chronic Lower Respiratory Disease: 0.2%
-   Stroke:0.2%

3.  Florida

-   Cancer: 1.4%
-   Heart Disease: 1.0%
-   Unintentional Injury: 0.5%
-   Chronic Lower Respiratory Disease:0.2%
-   Stroke: 0.2%

4.New York:

-   Cancer: 1.3%
-   Heart Disease: 0.9%
-   Unintentional Injury: 0.3%
-   Chronic Lower Respiratory Disease: 0.2%
-   Stroke: 0.1%

5.  Pennsylvania

-   Cancer: 1.0%
-   Heart Disease: 0.7%
-   Unintentional Injury: 0.4%
-   Chronic Lower Respiratory Disease: 0.1%
-   Stroke: 0.1%

6.  Ohio

-   Cancer: 0.9%
-   Heart Disease: 0.7%
-   Unintentional Injury: 0.3%
-   Chronic Lower Respiratory Disease: 0.2%
-   Stroke: 0.1%

7.  Illinois

-   Cancer: 0.9%
-   Heart Disease: 0.6%
-   Unintentional Injury: 0.3%
-   Chronic Lower Respiratory Disease: 0.1%
-   Stroke: 0.1%

8.  Michigan

-   Cancer: 0.7%
-   Heart Disease: 0.6%
-   Unintentional Injury: 0.2%
-   Chronic Lower Respiratory Disease: 0.1%
-   Stroke: 0.1%

9.  North Carolina

-   Cancer: 0.7%
-   Heart Disease: 0.5%
-   Unintentional Injury: 0.3%
-   Chronic Lower Respiratory Disease: 0.1%
-   Stroke: 0.1%

10. Georgia

-   Cancer: 0.6%
-   Heart Disease: 0.5%
-   Unintentional Injury: 0.3%
-   Chronic Lower Respiratory Disease: 0.1%
-   Stroke: 0.1%

## Conclusion:

The analysis of data from 2005 to 2015 reveals that Cancer is the leading cause of death, steadily rising over the years. Heart Disease follows, showing a slight decrease until 2013, and then a slight increase. Unintentional Injuries are continuously rising, presenting a growing public health concern. State-wise, California, Texas, and Florida report the highest number of deaths, with Cancer and Heart Disease consistently ranking at the top across all states analyzed.

Although deaths from Chronic Lower Respiratory Disease and Stroke remained stable from 2005 to 2015, any small increases in these areas should be watched closely to prevent them from becoming bigger health problems. The steady rise in deaths from Unintentional Injuries shows a need for targeted public health efforts and preventive measures to reduce these incidents. The state-by-state analysis reveals that Cancer and Heart Disease are the leading causes of death in all states, but the differences between states might be influenced by other factors. The visualizations and summary statistics highlight important trends and patterns and are valuable tools for policymakers.

Future research could look into underlying factors like demographics, policies, and access to affordable healthcare behind the trends seen in Cancer, Heart Disease, and Unintentional Injuries. A deeper understanding of these factors can lead to better health policies and interventions.
